Estimating dark matter halo concentration using the integrated mass profile

### 2015-12-24 paper/paper.tex
fixed typos

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it realiable in cases when low numerical resolution becomes a
limitation for other methods.   
Additionally, the algorithm is straightforward to implement.  
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of our method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
For halos sampled with $20$ particles our method recovers the input
concentration with $10\%$ accuracy, while for the maximum radial
velocity method it is on the order of $20\%$ while density profile
methods just fail at such low resolution limits.
For halos with $2\times 10^4$ particles our method achieves a recovery
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{12}\hMsun <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We only find a small deviation at low masses, $M<10^{13}\hMsun$, where
our method yields lower median concentration values by $20\%-30\%$
compared to the velocity and density methods. 
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

The internal DM distribution in a halo can be parameterized through the density
profile.  
In a first approximation this profile is spherically symmetric and
its density only dependens on the radial coordinate. 
One the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile 
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution are still 
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

@@

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However, it effectively takes into account only a single data point and
discards the behaviour of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  
Small fluctuations on the value of this maximum can yield large
perturbations on the estimated concentration parameter.

@@

Our method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).
We build it from N-body data following the next steps.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. 
In this process we discard the particle at the center.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. Our method provide the most accurate estimate at fixed
  particle number $N$.  
    \label{fig:error}}
\end{figure}

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
  \end{center}
  \caption{Comparison between the concentrations measured by our
    method and the maximum velocity (left) and density (middle)
    methods. The solid line indicates the equal value between the two
    techniques. The dashed line shows the upper 15 per cent difference
    between two methods.
  \label{fig:comparison}}
\end{figure*}

@@

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by 
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$. 
A large discrepacy between the estimated and input concentrations are
quantified by a large $\avg{|D|}$.

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{concentration.pdf}
\end{center}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using our method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}}
\end{figure}

@@

At fixed particle numbers our method always shows the lowest
$\avg{|D|}$ values compared to the other two.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
decreasing to $0.01\%$ for halos with $20000$ particles.  
The dependence of $\avg{|D|}$ with the particle number $N$
goes approximately as $\avg{|D|}\propto N^{-1/2}$.   
This hints that the increasing accuracy of the method could be due to
a decrease of Poisson noise.

The method based on the maximum of the circular velocity shows a
similar behaviour with $\avg{|D|}\propto N^{-1/3}$.  
Its accuracy is $2-5$ times lower than in our method, on the order of
$20\%$ for $20$ particle halos and $0.5\%$ for $20000$ particle
halos. 
The method  based on the direct density fit shows the lowest accuracy for a low
  particle number. 
  As it is discussed in \citep{Munoz2011}, density binning for halos
  with particle numbers below $\sim200$ leads to a biased estimation
  of the mass density profile, and therefore to a biased estimation of
  the concentration parameter. 
This behaviour is evident in the large values of $\avg{|D|}$ for halos
with number of particles below $\sim 200$ and an intermediate accuracy
between the other two methods for a high particle number.

@@

The results presented in the previous section show that in an
idealized setup (pure NFW desity profiles, perfect spherical halos in
total isolation) our method performs better than the other two
methods commonly used in the literature.

However, simulated dark matter halos in a cosmological context are
complex. 
They are not isolated, many of them  experience mergers that take them
out of the equilibrium. 
Halos may have also plenty of substructure. All these effects
disturb the mass density profile, taking it  away from the ideal NFW
profile. 
In order to explore the performance of our method in this setting, we
use it to estimate the  concentration parameter in dark matter halos
obtained from a cosmological N-body simulation and compare with the
other two classical methods.

@@

%%%%%%%

@@

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ (corresponding to $200$
times the critical density) to select an spherical region that we
redefine to be our halo.  
This choice makes that our overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density method we
only report results from overdensities with at least $200$ particles.

Figure \ref{fig:comparison} shows the comparison between the
concentration values obtained by the three different methods. 
The diagonal line marks a one-to-one correspondence if all the values
were equal.
The three methods show a broad general agreement. 
The largest difference is present at high concentration $c>5$ where
our method produces systematic lower values than the other two
methods. Most of the differences are at the $20\%$ level, with some
extreme cases for the highest concentration values with $50\%$
differences.  The velocity and density methods produce consistent
values within a $20\%$ scatter, only some halos with very high
concentrations $c\sim 10$ have a $15\%$ lower concentration with the
velocity method.

We explored halo shape and relaxation as a source of this
differences. We found shape does not bias the concentration
estimation; relaxation, quantified by the
offset between the center of mass and minimum of the potential, can
only explain at most $5\%$ of the offset between methods.  We note
that the highest concentration values correspond to the smallest halos
with $\sim 200$ particles, a regime where both the density and
velocity methods are reliable only the $5\%$ in a perfect spherical
halo.  
This difference might actually signal a deviation from the NFW profile
due to the low resolution in the simulation. From this we suggest that
asserting with confidence the systematic difference between the
different methods should be confirmed  a higher resolution simulation
and larger statistics. We leave this comparison for future work.

Figure \ref{fig:concentration} summarizes the previous results in the
mass-concentration relationship. 
We find that the median concentraion
can be fit by the following relationship
\begin{equation} c_{200} = 5.62 \left(\frac{M_{200}}{10^{12}\hMsun}\right)^{-0.061}, 
\end{equation}
%
which has a similar power value compared to the $-0.075$ reported by
\citep{Prada2012}, while the normalization is $\sim 20\%$ lower
compared to the $7.28$ value reported by the same authors.

@@

In this letter we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  
We tested the method on mock halo data to study the impact of total
number of particles and input concentration on the retrieved values.  
We compared these results against two other methods commonly used in
the literature to estimate concentrations.  
For all methods, the accuracy in retrieving the input concentration
increases with the number of particles as summarized in Figure
\ref{fig:error}.  Our method systematically overperforms the other two.

We also applied the method to halos extracted from a cosmological
N-body simulation to estimate the impact on the mass concentration
relationship. Although the three  methods are in broad agreement there
are some noticeable differences in the mass concentration relationship
at lower masses $<10^{13}\hMsun$ at the $20-30\% level$.

@@

 The authors acknowldge the technical support from the new
 high-performance computing facility at Uniandes. JEF-R acknowledges
 financial support from Vicerrector\'ia de Investigaciones at Uniandes
 through a FAPA project. JCMC acknowledges financial support from
 ``Estrategia de  sostenibilidad 2014-2015, Universidad de
 Antioquia''.

>>>>

\begin{abstract}
We present a new algorithm to estimate the concentration of N-body
dark matter halos using the integrated mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.   
Additionally, the algorithm is straightforward to implement.  
We test the performance of this method by estimating halo
concentration both on mock and N-body halos. 
We compare these results against two other methods:
maximum radial velocity measurements and radial particle binning.
Tests on the mock halos show that the accuracy of the new method to
recover known input concentrations varies with halo resolution,
outperforming the other two methods. 
For halos sampled with $20$ particles the new method recovers the input
concentration with $10\%$ accuracy, while for the maximum radial
velocity method it is on the order of $20\%$ while density profile
methods just fail at such low resolution limits.
For halos with $2\times 10^4$ particles the new method achieves a recovery
accuracy of $0.01\%$ while the velocity and density profile methods
achieve $0.1\%$ and $1\%$, respectively. 
We also measure the mass-concentration relationship on N-body
data. 
We find that in the probed mass range ($10^{12}\hMsun <
10^{14}\hMsun$) the three methods give consistent results within the
statistical uncertainties.   
We only find a small deviation at low masses, $M<10^{13}\hMsun$, where
the new method yields lower median concentration values by $20\%-30\%$
compared to the velocity and density methods. 
From these results we believe that the new method is a promising tool
to probe the internal structure of dark matter halos.
\end{abstract}

@@

The internal DM distribution in a halo can be parameterized through the density
profile.  
In a first approximation this profile is spherically symmetric and
its density only depends on the radial coordinate. 
One the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile 
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution are still 
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

@@

The second method solves the two problems mentioned above.  
It works with low particle numbers and does not involve data binning.  
However, it effectively takes into account only a single data point and
discards the behavior of the ratio $V_{\rm circ}/V_{\rm vir}$ below
and above its maxima.  
Small fluctuations on the value of this maximum can yield large
perturbations on the estimated concentration parameter.

@@

The new method uses the integrated mass profile defined in
Eq. (\ref{eq:profile}).
We build it from N-body data following the next steps.
First, we define the center of the halo to be at the position of the
particle with the lowest gravitational potential.  
Then we rank the particles by their increasing radial distance from
the center. 
From this ranked list of $i=1,N$ particles, the total mass at a radius
$r_i$ is $M_i=i\times m_p$, where $r_i$ is the position of the $i$-th
particle and $m_p$ is the mass of a single computational particle. 
In this process we discard the particle at the center.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.  
    \label{fig:error}}
\end{figure}

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
  \end{center}
\vspace{-0.5cm}
  \caption{Comparison between the concentrations measured by the new 
    method and the maximum velocity (left) and density (middle)
    methods. The solid line indicates the equal value between the two
    techniques. The dashed line shows the upper 15 per cent difference
    between two methods.
  \label{fig:comparison}}
\end{figure*}

@@

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by 
%
\begin{equation}
D=(c_{in}-c_{out})/c_{in},
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$. 
A large discrepancy between the estimated and input concentrations are
quantified by a large $\avg{|D|}$.

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{concentration.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using the new method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}}
\end{figure}

@@

At fixed particle numbers the new method always shows the lowest
$\avg{|D|}$ values compared to the other two.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
decreasing to $0.01\%$ for halos with $20000$ particles.  
The dependence of $\avg{|D|}$ with the particle number $N$
goes approximately as $\avg{|D|}\propto N^{-1/2}$.   
This hints that the increasing accuracy of the method could be due to
a decrease of Poisson noise.

The method based on the maximum of the circular velocity shows a
similar behaviour with $\avg{|D|}\propto N^{-1/3}$.  
Its accuracy is $2-5$ times lower than in the new method, on the order of
$20\%$ for $20$ particle halos and $0.5\%$ for $20000$ particle
halos. 
The method  based on the direct density fit shows the lowest accuracy for a low
  particle number. 
  As it is discussed in \citep{Munoz2011}, density binning for halos
  with particle numbers below $\sim200$ leads to a biased estimation
  of the mass density profile, and therefore to a biased estimation of
  the concentration parameter. 
This behaviour is evident in the large values of $\avg{|D|}$ for halos
with number of particles below $\sim 200$ and an intermediate accuracy
between the other two methods for a high particle number.

@@

The results presented in the previous section show that in an
idealized setup (pure NFW density profiles, perfect spherical halos in
total isolation) the new method performs better than the other two
methods commonly used in the literature.

However, simulated dark matter halos in a cosmological context are
complex. 
They are not isolated, many of them  experience mergers that take them
out of the equilibrium. 
Halos may have also plenty of substructure. 
All these effects disturb the mass density profile, taking it  away
from the ideal NFW profile. 
In order to explore the performance of the new method in this setting, we
use it to estimate the  concentration parameter in dark matter halos
obtained from a cosmological N-body simulation and compare it with the
other two classical methods.

@@

@@

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ (corresponding to $200$
times the critical density) to select an spherical region that we
redefine to be our halo.  
This choice makes that the overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density method we
only report results from overdensities with at least $200$ particles.

Figure \ref{fig:comparison} shows the comparison between the
concentration values obtained by the three different methods. 
The diagonal line marks a one-to-one correspondence if all the values
were equal.
The three methods show a broad general agreement. 
The largest difference is present at high concentration $c>5$ where
the new method produces systematic lower values than the other two
methods. Most of the differences are at the $20\%$ level, with some
extreme cases for the highest concentration values with $50\%$
differences.  The velocity and density methods produce consistent
values within a $20\%$ scatter, only some halos with very high
concentrations $c\sim 10$ have a $15\%$ lower concentration with the
velocity method.

We explored halo shape and relaxation as a source of this
differences. We found that shape does not bias the concentration
estimation; relaxation, quantified by the
offset between the center of mass and minimum of the potential, can
only offset at most $5\%$ the median results between methods.  
We note that the highest concentration values correspond to the smallest halos
with $\sim 200$ particles, a regime where both the density and
velocity methods are reliable only the $5\%$ in a perfect spherical
halo.

The differences we find might signal a deviation from the NFW profile
due to the low resolution of these halos. We suggest that asserting
with confidence the systematic difference between the different
methods should be confirmed  a higher resolution simulation and larger
statistics. We leave this comparison for future work.

Figure \ref{fig:concentration} summarizes the previous results in the
mass-concentration relationship. 
We find that the median concentration
can be fit by the following relationship
\begin{equation} c_{200} = 5.62 \left(\frac{M_{200}}{10^{12}\hMsun}\right)^{-0.061}, 
\end{equation}
%
which has a similar power value compared to the $-0.075$ reported by
\citep{Prada2012}, while the normalization is $\sim 20\%$ lower
compared to the $7.28$ value reported by the same authors.

@@

In this letter we presented a new method to estimate the concentration
of dark matter halos in N-body simulations.  
We tested the method on mock halo data to study the impact of total
number of particles and input concentration on the retrieved values.  
We compared these results against two other methods commonly used in
the literature to estimate concentrations.  
For all methods, the accuracy in retrieving the input concentration
increases with the number of particles as summarized in Figure
\ref{fig:error}.  The new method systematically outperforms the other two.

We also applied the method to halos extracted from a cosmological
N-body simulation to estimate the impact on the mass concentration
relationship. Although the three  methods are in broad agreement there
are some noticeable differences in the mass concentration relationship
at lower masses, $<10^{13}\hMsun$, at the $20-30\%$ level.

@@

 The authors acknowledge the technical support from the new
 high-performance computing facility at Uniandes. JEF-R acknowledges
 financial support from Vicerrector\'ia de Investigaciones at Uniandes
 through a FAPA project. JCMC acknowledges financial support from
 ``Estrategia de  sostenibilidad 2014-2015, Universidad de
 Antioquia''.
